---
title: "Dietary polyphenol and obesity markers. Translational research"
author: "Izar de Villasante, Brainvitge"
date: '`r format(Sys.Date(),"%e de %B, %Y")`'
#runtime: shiny
output:
  html_document:
    df_print: paged
    toc: yes
    toc_depth: 3
  word_document:
    toc: yes
    toc_depth: '3'
  pdf_document:
    keep_tex: yes
    toc: yes
    toc_depth: 3
params:
  
  
  
  
  main_folder: !r getwd()
  #/home/izar/notebooks/
  
  men_color: "aquamarine"  #hcl(h=60,c=100,l=65) 
  women_color: "indianred1"      #hcl(h=15,c=100,l=65) #"#ED813E" # 
  results: ./Results
  Datasets: ./Datasets
  mercedes: "EPIC.PANACEA_log2_factors" 
  excel_local: "Local_Measures.xlsx"
  variables: [vol]
    
header-includes: 
  - \usepackage{bbm}
  - \usepackage[spanish]{babel}
---
\newline
\newline

---

```{r setup, include=FALSE}
require(knitr)
#rmarkdown::render("panacea.Rmd", params = "ask")

# include this code chunk as-is to set options
opts_chunk$set(comment = NA, prompt = TRUE, tidy = FALSE, 
               fig.width = 10, fig.height = 10,echo = TRUE, 
               message = FALSE, warning = FALSE, cache=FALSE)
Sys.setlocale("LC_TIME", "C")
```

```{r paquetes, include=FALSE}
if(!(require(faraway))) install.packages("faraway")
if(!(require(devtools))) install.packages("devtools")
if(!(require(lmerTest))) install.packages("lmerTest")
if(!(require(ggplot2))) install.packages("ggplot2")
if(!(require(ggpubr)))install.packages("ggpubr")
if(!(require(grid)))install.packages("grid")
if(!(require(gtable)))install.packages("gtable")
if(!(require(magrittr)))install.packages("magrittr")

if(!(require(dplyr)))install.packages("dplyr")
 
if(!(require(printr))) {
  install.packages(
    'printr',
    type = 'source',
    repos = c('http://yihui.name/xran', 'http://cran.rstudio.com')
  )
}
```

# Contexto del Proyecto:

El objetivo principal del estudio es evaluar la asociación entre la exposición de polifenoles y los marcadores de obesidad e investigar su efecto sobre los parámetros de salud y la pérdida ponderal en un tratamiento de adelgazamiento. El proyecto consta de tres fases: i) evaluación de la exposición de polifenoles, medidos mediante cuestionarios de dieta, biomarcadores o la combinación de ambas medidas, y el cambio de peso y perímetro de cintura en un gran estudio epidemiológico observacional: EPIC-PANACEA, que incluye ~370000 adultos provenientes de 10 países europeos, con un seguimiento de 2 a 11 años.

*The main objective is to evaluate the association between polyphenol exposure and obesity markers and to investigate their effect on several health markers and the body weight composition in a weight loss treatment. The Project is divided in 3 stages: i) Firstly, to evaluate the association between the exposure of polyphenols, assessed using dietary questionnaires, biomarkers or the combination of both methods, and the change in body weight and waist circumference in a large observational study: the EPIC-PANACEA. It consists in ~370000 adult men and women from ten European countries with a follow-up ranged from 2 to 11 years.*


```{r load data}
#First save current working directory:
cwd <- getwd()
#Load Datasets folder:
datasets <- paste(cwd,"Datasets/",sep="/")

#Load each file:

file1 <- "bdpanacea_reduced.txt"
file2 <- "EPIC.PANACEA_log2_factors.txt"



#reduced <- read.table("Z:/Rstudio/EPIC/bdpanacea_reduced.txt", quote="\"", comment.char="")
if(!exists("datos")){
  datos <- read.table(paste0(datasets,file2), quote="\"", comment.char="")
}
```

```{r}

datos[12:21]<-lapply(datos[12:21],function(x) {as.factor(x)})

levels(datos$Sex_f)<-c("Male","Female")
str(datos$Sex_f)

levels(datos$Country_f) <- c("France","Italy","Spain","UK","Netherlands","Greece","Germany","Sweden","Denmark","Norway")
str(datos$Country_f)
str(datos, list.len=30)


```

```{r}
library(leaps)
datos_main <- datos[,c("Age_Recr","Weight_change_5y", "Country_f","Sex_f")]
datos_poli<-datos[,23:ncol(datos)]
sub_datos <- cbind(datos_main,datos_poli)
#b <- regsubsets(Weight_change_5y ~ ., data =sub_datos,really.big = T )

```

```{r}
#rs <- summary(b)
#rs$which

#plot(1:9,rs$adjr2,xlab="No. of Parameters",ylab="Adjusted R-square")

```

```{r}
#best_sub <- which.max(rs$adjr2)
#which(rs$which[best_sub,])
```

There is a wise option which would be to make a best subsets forcing all covariates (Region, sex, age) to be included. in this case we will set the maximum size of the model to 12:

```{r}
#b2 <- regsubsets(Weight_change_5y ~ ., 
#                 data =sub_datos,
#                 really.big = T ,
#                 force.in = c(1,3,4),
#                 nvmax=12)

b2 <- 
```
```{r}
rs <- summary(b2)
rs$which
plot(1:length(rs$adjr2),rs$adjr2,xlab="No. of Parameters",ylab="Adjusted R-square")

```




```{r}
require(magrittr)
require(dplyr)
r2_out <- rs$outmat[,which(apply(rs$outmat,2,function(r) any(r %in% "*")))]

```



```{r}

best_sub <- which.max(rs$adjr2)
y<-rs$which[best_sub,]
rs$outmat[,which(y)]

```

```{r}
saveRDS(b2, file="Bestsubsets12_with_covariates.RDS")
```


