---
title: "Dietary polyphenol and obesity markers. Translational research"
author: "Izar de Villasante, Brainvitge"
date: '`r format(Sys.Date(),"%e de %B, %Y")`'
#runtime: shiny
output:
  html_document:
    df_print: paged
    toc: yes
    toc_depth: 3
  word_document:
    toc: yes
    toc_depth: '3'
  pdf_document:
    keep_tex: yes
    toc: yes
    toc_depth: 3
params:
  
  
  
  
  main_folder: !r getwd()
  #/home/izar/notebooks/
  
  men_color: "aquamarine"  #hcl(h=60,c=100,l=65) 
  women_color: "indianred1"      #hcl(h=15,c=100,l=65) #"#ED813E" # 
  results: ./Results
  Datasets: ./Datasets
  mercedes: "EPIC.PANACEA_log2_factors" 
  excel_local: "Local_Measures.xlsx"
  variables: [vol]
    
header-includes: 
  - \usepackage{bbm}
  - \usepackage[spanish]{babel}
---
\newline
\newline

---

```{r setup, include=FALSE}
require(knitr)
#rmarkdown::render("panacea.Rmd", params = "ask")

# include this code chunk as-is to set options
opts_chunk$set(comment = NA, prompt = TRUE, tidy = FALSE, 
               fig.width = 10, fig.height = 10,echo = TRUE, 
               message = FALSE, warning = FALSE, cache=FALSE)
Sys.setlocale("LC_TIME", "C")
```

```{r paquetes, include=FALSE}
if(!(require(faraway))) install.packages("faraway")
if(!(require(devtools))) install.packages("devtools")
if(!(require(lmerTest))) install.packages("lmerTest")
if(!(require(ggplot2))) install.packages("ggplot2")
if(!(require(ggpubr)))install.packages("ggpubr")
if(!(require(grid)))install.packages("grid")
if(!(require(gtable)))install.packages("gtable")
if(!(require(magrittr)))install.packages("magrittr")

if(!(require(dplyr)))install.packages("dplyr")
 
if(!(require(printr))) {
  install.packages(
    'printr',
    type = 'source',
    repos = c('http://yihui.name/xran', 'http://cran.rstudio.com')
  )
}
```

# Contexto del Proyecto:

El objetivo principal del estudio es evaluar la asociación entre la exposición de polifenoles y los marcadores de obesidad e investigar su efecto sobre los parámetros de salud y la pérdida ponderal en un tratamiento de adelgazamiento. El proyecto consta de tres fases: i) evaluación de la exposición de polifenoles, medidos mediante cuestionarios de dieta, biomarcadores o la combinación de ambas medidas, y el cambio de peso y perímetro de cintura en un gran estudio epidemiológico observacional: EPIC-PANACEA, que incluye ~370000 adultos provenientes de 10 países europeos, con un seguimiento de 2 a 11 años.

*The main objective is to evaluate the association between polyphenol exposure and obesity markers and to investigate their effect on several health markers and the body weight composition in a weight loss treatment. The Project is divided in 3 stages: i) Firstly, to evaluate the association between the exposure of polyphenols, assessed using dietary questionnaires, biomarkers or the combination of both methods, and the change in body weight and waist circumference in a large observational study: the EPIC-PANACEA. It consists in ~370000 adult men and women from ten European countries with a follow-up ranged from 2 to 11 years.*




In this case a best subsets including the 3 covariates Age, Sex and Region has been performed up to 12 predictors.

```{r load data}
datos <- readRDS("EPIC.PANACEA_log2_factors.RDS")
subsets <- readRDS("Bestsubsets12_with_covariates.RDS")

```


```{r}
library(leaps)
rs <- summary(subsets)
plot(4:13,rs$adjr2,xlab="No. of Parameters",ylab="Adjusted R-square")

```




```{r}
require(magrittr)
require(dplyr)
r2_out <- rs$outmat[,which(apply(rs$outmat,2,function(r) any(r %in% "*")))]
r2_out <- cbind(r2_out,r2=round(rs$adjr2,4))
r2_out
```


```{r}
coef(subsets,which.max(rs$adjr2))
paste0("Best model R2:", max(rs$adjr2))

```


Another approach is to try a regression without any covariate. Only polifenols:

```{r}
subsets <- readRDS("Bestsubsets10_polifenol_only.RDS")
```



```{r}
library(leaps)
rs <- summary(subsets)
plot(4:13,rs$adjr2,xlab="No. of Parameters",ylab="Adjusted R-square")

```




```{r}
require(magrittr)
require(dplyr)
r2_out <- rs$outmat[,which(apply(rs$outmat,2,function(r) any(r %in% "*")))]
r2_out <- cbind(r2_out,r2=round(rs$adjr2,4))
r2_out
```


```{r}
coef(subsets,which.max(rs$adjr2))
paste0("Best model R2:", max(rs$adjr2))

```

We can observe some interesting synergies here. Although the best predictor alone is pp638 it doesn't work as well in combination to others. Imediately after the single polifenol, the combo pp639+pp642+pp644 has a good predictive power explaining up to 2,48% of the variance in weight.   Also we see that including pp561 improves the adjustment more than adding pp555 or pp357 alone but less than adding them together. 

I would take a closer look to the model in the 3rd round:
```{r}
coef(subsets,3)

```
```{r}
polifenol_simple_model <- lm(data=datos,Weight_change_5y ~ pp639+pp642+pp644)
```

```{r}
mixed_model <- lme4::lmer(data=datos,Weight_change_5y ~ pp639+pp642+pp644+ Sex_f + Centre_f + Age_Recr+(1|Centre_f))
summary(mixed_model)
```

```{r}
mixed_model_random <- lme4::lmer(data=datos,Weight_change_5y ~ pp639+pp642+pp644+ Sex_f + Centre_f + Age_Recr+(1+pp639+pp642+pp644|Centre_f))
summary(mixed_model)
```


```{r}
AIC(polifenol_simple_model,mixed_model,mixed_model_random)
```

```{r}
anova(mixed_model,mixed_model_random)

```
