---
title: "Dietary polyphenol and obesity markers. Translational research"
author: "Izar de Villasante, Brainvitge"
date: '`r format(Sys.Date(),"%e de %B, %Y")`'
#runtime: shiny
output:
  html_document:
    df_print: paged
    toc: yes
    toc_depth: 3
  word_document:
    toc: yes
    toc_depth: '3'
  pdf_document:
    keep_tex: yes
    toc: yes
    toc_depth: 3
params:
  
  
  
  
  main_folder: !r getwd()
  #/home/izar/notebooks/
  
  men_color: "aquamarine"  #hcl(h=60,c=100,l=65) 
  women_color: "indianred1"      #hcl(h=15,c=100,l=65) #"#ED813E" # 
  results: ./Results
  Datasets: ./Datasets
  mercedes: "EPIC.PANACEA_log2_factors" 
  excel_local: "Local_Measures.xlsx"
  variables: [vol]
    
header-includes: 
  - \usepackage{bbm}
  - \usepackage[spanish]{babel}
---
\newline
\newline

---

```{r setup, include=FALSE}
require(knitr)
#rmarkdown::render("panacea.Rmd", params = "ask")  #Use in case you want to manually select the file to load data from.

# include this code chunk as-is to set options
opts_chunk$set(comment = NA, prompt = TRUE, tidy = FALSE, 
               fig.width = 10, fig.height = 10,echo = TRUE, 
               message = FALSE, warning = FALSE, cache=TRUE)
Sys.setlocale("LC_TIME", "C")
```

```{r paquetes, include=FALSE}


#Install & Load packages for plots:
if(!(require(ggplot2))) install.packages("ggplot2")
if(!(require(ggpubr)))install.packages("ggpubr")
if(!(require(grid)))install.packages("grid")
if(!(require(gtable)))install.packages("gtable")
if(!(require(gtable)))install.packages("plot3D")

#Install & Load packages for multiple regression:
if(!(require(lme4)))install.packages("lme4")
if(!(require(lmerTest)))install.packages("lmerTest")
if(!(require(extraoperators)))install.packages("extraoperators")
if(!(require(multilevelTools)))install.packages("multilevelTools")

#Install devtools with "make" in order to compile packages from binaries, ex: from github.
if(!(require(devtools))) install.packages("devtools")

#Example of package installation specifying repository:
if(!(require(printr))) {
  install.packages(
    'printr',
    type = 'source',
    repos = c('http://yihui.name/xran', 'http://cran.rstudio.com')
  )
devtools::install_github("AckerDWM/gg3D")

library("gg3D")
}
```

# Contexto del Proyecto:

El objetivo principal del estudio es evaluar la asociación entre la exposición de polifenoles y los marcadores de obesidad e investigar su efecto sobre los parámetros de salud y la pérdida ponderal en un tratamiento de adelgazamiento. El proyecto consta de tres fases: i) evaluación de la exposición de polifenoles, medidos mediante cuestionarios de dieta, biomarcadores o la combinación de ambas medidas, y el cambio de peso y perímetro de cintura en un gran estudio epidemiológico observacional: EPIC-PANACEA, que incluye ~370000 adultos provenientes de 10 países europeos, con un seguimiento de 2 a 11 años.

*The main objective is to evaluate the association between polyphenol exposure and obesity markers and to investigate their effect on several health markers and the body weight composition in a weight loss treatment. The Project is divided in 3 stages: i) Firstly, to evaluate the association between the exposure of polyphenols, assessed using dietary questionnaires, biomarkers or the combination of both methods, and the change in body weight and waist circumference in a large observational study: the EPIC-PANACEA. It consists in ~370000 adult men and women from ten European countries with a follow-up ranged from 2 to 11 years.*


**¡¡¡¡Important note: Only 9 countries participate, after Greece quitting the project, the data we will use comes from a curated dataset provided by Mercedes Gil Lespinard with some modifications. The most relevant being the addition of the variable containing the total amount of polifenol which is log2 transformed to remove outliers. We will use this variable along the report !!!!**



```{r load data }
#First save current working directory:
cwd <- getwd()
#Load Datasets folder:
datasets <- paste(cwd,"Datasets/",sep="/")

#Load each file:

file1 <- "bdpanacea_reduced.txt"
file2 <- "EPIC.PANACEA_log2_factors.txt"



#reduced <- read.table("Z:/Rstudio/EPIC/bdpanacea_reduced.txt", quote="\"", comment.char="")
if(!exists("datos")){
  datos <- read.table(paste0(datasets,file2), quote="\"", comment.char="")
}
```

```{r}
str(datos, list.len=30)
```
```{r}
datos[12:21]<-lapply(datos[12:21],function(x) {as.factor(x)})
str(datos, list.len=30)
```



# Data Visualization
To take a first glance to our data, the first step is to generate some plots:

## Confounding variables

Technical and biological annotations are often interrelated, leading to confounding.
Because of the nature of the project, some variables were covaried along with polifenol intake in order to diminish the effect of external variability to our model.
The variables Sex, Age and centre are proposed as confounding variables in our model and they will be covariated in order to improve accuracy.
Since each centre collected data in a singular way, some centre-specific variance may be introduced to our data. 

## Sex:
So let's start looking at the variable *"Sex_f"*.

In this case, we will treat biological sex as a binary variable with the values Male/Female 

```{r}
levels(datos$Sex_f)<-c("Male","Female")
str(datos$Sex_f)
levels(datos$Country_f) <- c("France","Italy","Spain","UK","Netherlands","Greece","Germany","Sweden","Denmark","Norway")
str(datos$Country_f)
```





```{r}
t_sex<-table(datos$Centre_f,datos$Sex_f)
t_sex <- cbind(t_sex, Total = rowSums(t_sex))   #Afegim columna de totals a la dreta
t_sex <- rbind(t_sex, Total = colSums(t_sex))   #Afegim fila de totals a sota.
t_sex

# No volem agafar la última dada corresponent al total per fer el calcul de mitja, mínim, màxim i desviació estandard.
# Per accedir a la ultima posició podem fer (nrow(t_sex))
# Per eliminar aquesta observació afegim un - devant.
# Totes les dades excepte la última fila    -->    t_sex[-nrow(tsex,)]

paste("Average amount of participants per centre: ",as.integer(mean(t_sex[-nrow(t_sex),3])))
paste("centre with greatest amount of participants: ", max(t_sex[-nrow(t_sex),3]))
paste("centre with fewest amount of participants: ", min(t_sex[-nrow(t_sex),3]))
paste("Standard deviation: ", sd(t_sex[-27,3]))


#which(t_sex[,3]==min(t_sex[,3])) #Gives the name of the Centre


```
There are huge differences between the number of participants from each centre. This reinforces discrimination by centre to avoid the effect of the different weights of centre-specific errors. In other words, centres with a lot of participants would impact in a bigger manner than centres with fewer participants . 

```{r}

# Simple Pie Chart
slices <- c(sum(t_sex[,1]>0),sum(t_sex[,1]==0),sum(t_sex[,2]==0))
lbls <- c("Men + Women", "Only Women", "Only Men")
pct <- round(slices/sum(slices)*100)
lbls <- paste(lbls, pct) # add percents to labels
lbls <- paste(lbls,"%",sep="") # ad % to labels

pie(slices, labels = lbls, col=rainbow(length(lbls)),main="Pie Chart of sex representation of each centre")



```

### Boxplot 
Now that the representation of each sex in the data has been observed, let's take a look on the impact of belonging to each sex to obesity.  As  mentioned above, obesity is assessed by measuring weight change in the 5 years after the start of the study. A good way of getting an idea about the variance of the weight between groups and sex is a multiple boxplot of each centre separated by sex.  
```{r}
ggplot(datos, aes(x=Centre_f, y=Weight_change_5y, fill=Sex_f)) + 
    geom_boxplot()

```
We can see a rather similar shape of the boxes for each centre but with some differences, maybe some subgroups could be deduced from the plot. Maybe, this subgroups respond to different countries:

```{r}
ggplot(datos, aes(x=Centre_f, y=Weight_change_5y, fill=Country_f)) + 
    geom_boxplot()

```

Except for countries 7 and 8, it seems that the boxes of each country are quite similar. Nevertheless, if precission is what concerns us most, using centre variable insted of country will allways explain more variability, thus improving performance.  


There are a lot of outliers but they are symetrical , as well as a centered mean arround 0. This suggests a normal distribution of weight in the different centres. There doesn't appear to be great differences in weight loss/gain between sexs, since the quantile boxes present very similar structures  and means. 
In my opinion the most interesting fact we can extract from this plot is that the mean values seem to be a little bit displaced over or below 0 for each centre. This could be explained by center specific bias, maybe introduced by scale calibration errors, that may produce this fluctuations between centres. Of course, there are many other explanations, such as cultural tendency to lie about personal sensible data such as weight or other factors concerning the specific workflow of each Centre / Country.

In any case this variance suggests using a *mixed multilevel linear regression with fixed effect on region*.

**A check on the real effect of sex to our data shall be statistically assessed with a chi-squared test comparing the model precision with or without sex.**   

 

## Age:


The second variable proposed as counfounder is Age. 
Weight loss/gain is often related to specific stages of the life cycle. For instance, women tend to experience weight gain when entering the menopause. Also weight losses are common when approaching the final stages of life cycle.
Therefore, I propose to treat age as a Categorical variable.

```{r}
hist(datos$Age_Recr)
```
The distribution of the population is concentrated between 40 and 65 years. I would suggest a first aproximantion with the following groups:

* Under 40
* Between 40 and 50
* Between 50 and 60
* Between 60 and 70
* Above 70

```{r}
datos$Age <- cut_interval(datos$Age_Recr,n=5 ) #Creates sequence from 
levels(datos$Age) #

head(datos$Age)
#groups <- factor(datos$Age_Recr<=40, levels= c(0:40,41),labels=c("-40","40-50","50-60","60-70","+70"))
#groups
```

## Centre:

This may be the most difficult confounding variable to adress.
As previously proposed, there may be some variability introduced by the effects of events occurring on a specific centre. This centre-specific effects can either be fixed such as those observed in the weight bias of each centre or with rendom effects underlying the relationship between weight change and polifenol intake.

In order to asses this effect, the three following models with different degree of interaction in the Centre variable are proposed:

-Model1: Simple regression model with (Age, Sex and centre) as confounding variables. This model doesn't consider within-group variance in the centre variable. 

-Model2: Multilevel mixed model with fixed effect (different intercepts) for each centre  +Age +Sex. This model takes into account within-group variance in the centre variable, in the following way:
Imagine that the scale each centre uses to measure weights is biased. This would result in slightly different measures for each centre either higher than they should be or lower. Therefore, although the mean value (Intercept) may move, the relation between weight gain/loss and polifenol intake (Slope) should stay the same.  

-Model3: Multilevel mixed model with random effect for centre + Age + Sex. This model proposes that the relation between 5 year weight change and total polifenol intake may change between centers. Remind that we are talking about **TOTAL polifenol**, which is a composition of different groups of polifenols. The relation of each group of polifenols inside the class may vary from centre to centre, since the diet changes from one place to another. Thus, its effect on weight change. 

In first place, lets take a look to our data:


A scatter plot of 5y weight change vs total intake of polifeol for each centre is a good start:

```{r, fig.width=7,fig.height=70}

region_plot <- ggplot(aes(x=pp025,y=Weight_change_5y, colour=factor(Sex_f)),data=datos) + geom_point(size=0.2) + facet_grid(rows=vars(Centre_f),cols=vars(Sex_f),scales="free") + geom_smooth(method="lm",colour="black")
region_plot

```
There doesn't seem to exists a clear relation between polifenol intake and weight change. In general, a cloud of dots is drawn arround 0 value with no clear tendency. Nevertheless, a closer look should be taken to the line of dots near the -20 value that appears recursively. Centre 34 is a good example of this phenomenon.



```{r}
centre34 <- data.frame(datos[datos$Centre_f==34,c("pp025","Weight_change_5y","Sex_f")])
region_plot.34 <- ggplot(aes(x=pp025,y=Weight_change_5y, colour=factor(Sex_f)),data=centre34) + geom_point(size=2)  + geom_smooth(method="lm",colour="black")
region_plot.34

```
I would make the following educated guess about this phenomenon; the strange pattern observed in the left of the plot can be dur to:
*1. Data normalization: Data has been normalized, missing data or not consumers had a 0 value, which in a certain scale can result far away from standar deviation
- Solutions: If quantiles are used in a semi-continius way this is avoided. Also, adding a small value to all our data will correct this error, usually one magnitude order below the minimum. 

 **If it is because of missing values we should exclude this participants, since their exposure to polifenol is unknown and could be any value.** 
Another option is to separate 0 values as a control group.

# Models:

Just for instructive reasons, the needed libraries for the multilevel regressions and diagnostics will be loaded in the following chunk. Remember all the libraries where installed and loaded at the start of the report. 

```{r}
#The main package containing lmer function to call the regression:
library(lme4)  

#Packages for model diagnostics:
library(lmerTest)
library(extraoperators)
library(multilevelTools)
```

## Model 1: Simple linear regression:

This model will follow the structure of a simple regresion using the LSE:

y = 1 + x + C_i

* lmer: The call to the simple linear regression model formula.
* y: The dependent variable,  which we want to predict with our model.
* Pred: The predictors or covariates that we use in our model to predict y.
* 1: The Intercept. The value of y when predictors are 0. Also the position where the regression crosses or "Intercepts" the y-axis. There is no need to include it in our formula since it is already implicit.
* Cov_i: The other covariates that are not grouped.
* Grouping: The variable used to make grouping.

```{r}
Mod1 <- lm( Weight_change_5y ~ pp025 + Age_Recr + Sex_f + Centre_f,data = datos)
summary(Mod1)

```
```{r}
#Just to demonstrate that implicitly including the intercept term in our formula results in the same outcome:
Mod1_I<- lm( Weight_change_5y ~ 1 +pp025 + Age_Recr + Sex_f + Centre_f,data = datos)
summary(Mod1_I)

#WE can check in the following way:
coef(Mod1)==coef(Mod1_I)
```


The R^2 here is very low, this means that the accuracy of the model is terrible. 

In other words, the model fails to adjust propoerly to our data. A lot of error is present in the residuals which we can plot in the following way:

```{r}
Mod1_diagnostics <- plot(Mod1)
Mod1[1]
```



There could exist a problem of multicolineality between variables. This means that the different predicotr variables are intercorrelated, or that they explain the same variance:

```{r}
car::vif(Mod1)

```
A value of 10 or more tells us that multicollinearity exist. In that case a method to select variables shall be used. In this case, it seems that the poor performance of the model is not due to multicolinarity between variables. 

Just to clarify the concept of multicolinearity we will give the following example of intercorrelated variables:
If the variable country was added, one would expect to see collinearity with the Centre variable, since Centre is a subcategory inside Country, thus, the variability of each centre can be seen individually or combined inside each country. Let's check it out:

```{r}
Mod1_centre_country <- lm( Weight_change_5y ~ pp025 + Age_Recr + Sex_f + Centre_f + Country_f,data = datos)
#car::vif(Mod1_centre_country)
#<Error in vif.default(Mod1_centre_country) : there are aliased coefficients in the model

```
This means that there is perfect multicolinearity, which in turn means that two or more variables are linearly dependent. In other words, the variance in some variables is explained also by other variables. Let's check it:

```{r}
alias(Mod1_centre_country)
```
The output here is telling us that the variance explained by each country is the sum of variance explained by each centre. So, for 


```{r}
Mod1_country <- lm( Weight_change_5y ~ pp025 + Age_Recr + Sex_f +  Country_f,data = datos)
summary(Mod1_country)
```


## Model 2: Mixed linear regression with fixed effect:

```{r}
Mod2 <- lmer( Weight_change_5y ~ pp025 + Age_Recr + Sex_f + (1| Centre_f), datos)
summary(Mod2)


```


## Model 3: Mixed linear regression with random effect: 

The formula: `lmer(y ~ Pred + Cov_i + ( 1 + Pred | grouping)` Describes a model that corrects for random effects in the predictor variable. Let's stripe it down:

* lmer: The call to the multilevel regression formula provided by lme4 package.
* y: The dependent variable,  which we want to predict with our model.
* Pred: The predictors or covariates that we use in our model to predict y.
* 1: The Intercept. The value of y when predictors are 0. Also the position where the regression crosses or "Intercepts" the y-axis.
* Cov_i: The other covariates that are not grouped.
* Grouping: The variable used to make grouping.

As can be deduced from the above formula, the grouping ( 1 + Pred | grouping) affects both the intercept and the Predictor variable. Therefore, a new coefficient or slope is fitted for every group as well as the intercept. Remember that regression applyes the LSE function in order to minimize standard error produced by all the data points. In this case, every set of data inside grouping is evaluated separately, which results in different slopes for every group.




```{r}
Mod3 <- lmer( Weight_change_5y ~ pp025 + Age_Recr + Sex_f + (1 + pp025 | Centre_f), datos )
summary(Mod3)
```

```{r fig.width=7,fig.height=70}

region_plot_Mod1 <- ggplot(aes(x=pp025,y=Weight_change_5y, colour=factor(Sex_f)),data=datos) + geom_point(size=0.2) + facet_grid(rows=vars(Centre_f),cols=vars(Sex_f),scales="free") + geom_line(aes(y = predict(Mod1), colour= "var0")) +
geom_line(aes(y = predict(Mod2), colour= "var1")) +
  geom_line(aes(y = predict(Mod3), colour= "var2")) + geom_smooth(method="lm",colour="black")
region_plot_Mod1

```

Just to take a look to the data in a 3D fashion:
R repository don't include any library for 3D plotting but we can install 3rd-party libraries from github using Rtools:
```{r}
devtools::install_github("AckerDWM/gg3D")
library("gg3D")

```


```{r}

x <- datos$pp025
y <- datos$Weight_change_5y
z <- datos$Age_Recr
fit <- Mod1

# predict values on regular xy grid
grid.lines = 26
x.pred <- seq(min(datos$pp025), max(datos$pp025), length.out = grid.lines)
y.pred <- seq(min(datos$Weight_change_5y), max(datos$Age_Recr), length.out = grid.lines)
xyz <- expand.grid(pp025 = x.pred, Age_Recr = y.pred, Sex_f="Female",Centre_f="11")
z.pred <- matrix(predict(Mod1, newdata = xyz), nrow = grid.lines, ncol = grid.lines)
# fitted points for droplines to surface 
fitpoints <- predict(Mod1)

# scatter plot with regression plane
b <- plot3D::scatter3D( 
  x=datos$pp025,
  y=datos$Weight_change_5y,
  z=datos$Age_Recr,
  bty = "g",
  
  col.var = as.integer(datos$Sex_f),
  col = c(params$men_color, params$women_color),
  ticktype = "detailed",
  main ="3d",
  xlab = "polifenol", ylab = "Weight change", zlab = "Age",
  addlines = TRUE, length = 0.5, width = 0.5,
  labels = c("Male","Female") ,
  surf = list(x = x.pred, y = y.pred, z = z.pred,  facets = NA, fit = fitpoints) 
)
# scatter plot with regression plane
    #scatter3D(x, y, z, pch = 18, cex = 2, 
    #theta = 20, phi = 20, ticktype = "detailed",
    #xlab = "wt", ylab = "disp", zlab = "mpg",  
    #surf = list(x = x.pred, y = y.pred, z = z.pred,  
    #facets = NA, fit = fitpoints)
#)
b

```


```{r, fig.width=7,fig.height=70}

region_plot_Mod3_3d <- ggplot(data=datos, aes(x=pp025,y=Weight_change_5y, z=Age_Recr,colour=factor(Sex_f)))  + 
  axes_3D() + 
  stat_3D() +  
  facet_grid(rows=vars(Centre_f),cols=vars(Sex_f)) 



#region_plot_Mod3_3d <- ggplot(data=datos, aes(x=pp025,y=Weight_change_5y, z=Age_Recr,colour=factor(Sex_f)))  + axes_3D() + stat_3D() +  facet_grid(rows=vars(Centre_f),cols=vars(Sex_f),scales="free") + geom_line(aes(y = predict(Mod1), colour= "var0")) +geom_line(aes(y = predict(Mod2), colour= "var1")) +  geom_line(aes(y = predict(Mod3), colour= "var2"))  



#region_plot_3d <- ggplot(aes(x=pp025,y=Weight_change_5y,z=Age_f colour=factor(Sex_f)),data=datos) + geom_point(size=0.2) + facet_grid(rows=vars(Centre_f),cols=vars(Sex_f),scales="free") + geom_abline(aes(data.frame(coef$Mod1)))

region_plot_Mod3_3d
```



```{r}
#influencePlot(Mod1)
```


