---
title: "Dietary polyphenol and obesity markers. Translational research"
author: "Izar de Villasante, Brainvitge"
date: '`r format(Sys.Date(),"%e de %B, %Y")`'
#runtime: shiny
output:
  html_document:
    df_print: paged
    toc: yes
    toc_depth: 3
  word_document:
    toc: yes
    toc_depth: '3'
  pdf_document:
    keep_tex: yes
    toc: yes
    toc_depth: 3
params:
  
  
  
  
  main_folder: !r getwd()
  #/home/izar/notebooks/
  
  men_color: "aquamarine"  #hcl(h=60,c=100,l=65) 
  women_color: "indianred1"      #hcl(h=15,c=100,l=65) #"#ED813E" # 
  results: ./Results
  Datasets: ./Datasets
  mercedes: "EPIC.PANACEA_log2_factors" 
  excel_local: "Local_Measures.xlsx"
  variables: [vol]
    
header-includes: 
  - \usepackage{bbm}
  - \usepackage[spanish]{babel}
---
\newline
\newline

---

```{r setup, include=FALSE}
require(knitr)
#rmarkdown::render("panacea.Rmd", params = "ask")

# include this code chunk as-is to set options
opts_chunk$set(comment = NA, prompt = TRUE, tidy = FALSE, 
               fig.width = 10, fig.height = 10,echo = TRUE, 
               message = FALSE, warning = FALSE, cache=FALSE)
Sys.setlocale("LC_TIME", "C")
```

```{r paquetes, include=FALSE}
if(!(require(faraway))) install.packages("faraway")
if(!(require(devtools))) install.packages("devtools")
if(!(require(lmerTest))) install.packages("lmerTest")
if(!(require(ggplot2))) install.packages("ggplot2")
if(!(require(ggpubr)))install.packages("ggpubr")
if(!(require(grid)))install.packages("grid")
if(!(require(gtable)))install.packages("gtable")

if(!(require(gtools)))install.packages("gtools")
 
if(!(require(printr))) {
  install.packages(
    'printr',
    type = 'source',
    repos = c('http://yihui.name/xran', 'http://cran.rstudio.com')
  )
}
```

# Project context:

El objetivo principal del estudio es evaluar la asociación entre la exposición de polifenoles y los marcadores de obesidad e investigar su efecto sobre los parámetros de salud y la pérdida ponderal en un tratamiento de adelgazamiento. El proyecto consta de tres fases: i) evaluación de la exposición de polifenoles, medidos mediante cuestionarios de dieta, biomarcadores o la combinación de ambas medidas, y el cambio de peso y perímetro de cintura en un gran estudio epidemiológico observacional: EPIC-PANACEA, que incluye ~370000 adultos provenientes de 10 países europeos, con un seguimiento de 2 a 11 años.

*The main objective is to evaluate the association between polyphenol exposure and obesity markers and to investigate their effect on several health markers and the body weight composition in a weight loss treatment. The Project is divided in 3 stages: i) Firstly, to evaluate the association between the exposure of polyphenols, assessed using dietary questionnaires, biomarkers or the combination of both methods, and the change in body weight and waist circumference in a large observational study: the EPIC-PANACEA. It consists in ~370000 adult men and women from ten European countries with a follow-up ranged from 2 to 11 years.*

# Data Preparation:
## Load data

The first step is always to load the data. In this case the startingpoint is a database containing the polifenol intake from the Panacea project

```{r load data ,cache=False}
#First save current working directory:
cwd <- getwd()
#Load Datasets folder:
datasets <- paste(cwd,"Datasets/",sep="/")

#Load each file:

file1 <- "bdpanacea_reduced.txt"
file2 <- "weight.sas7bdat"



#reduced <- read.table("Z:/Rstudio/EPIC/bdpanacea_reduced.txt", quote="\"", comment.char="")
if(!exists("main_db")){
  main_db <- haven::read_sas(paste0(datasets,file2))
}
```


It is a very big dataset, let's take a quick look:
```{r}
dim(main_db)
pryr::object_size(main_db)
```
A total of 349165 participants from the "Epic-Panacea" project with consumption data of polifenols. A big dataset of 1.78GB
A total of 628 colums containing polifenol intake and other information that can be grouped as follows:

1. The first step is to convert the dataframe into a data.table. In order to make faster ops and easier to work. And most important save us RAM.
```{r}
library(data.table)
tbl<-setDT(main_db)
```


2. Now lets separate the columns containing general info such as Age, weight, etc. From the columns containing polifenols:

```{r}
general <- tbl[, .SD, .SDcols = c(1:3,490:555)]
str(general)

```

```{r}
sas.factor <- lapply(general,exists(attributes(X)$format.sas))
tst[,(cols):=lapply(.SD, as.factor),.SDcols=cols]
```




Let's have a look to the different variables.
```{r}
library(ggplot2)
p <- ggplot(melt(general), aes(factor(variable), value)) 
p + geom_boxplot() + facet_wrap(~variable, scale="free")
wa
```


```{r}
polifenols <- tbl[, .SD, .SDcols = 4:422]

```


3. [423:459] From column 423 to 459 we have 37 main subclasses of polifenols:

```{r}
names(main_db)[423:459]
PP_subclass <- names(main_db)[423:459]
#PP_subclass_db <- main_db[,423:459]
length(PP_subclass)
```

4. [460:497] From 460 to 487 there are 28 subclasses of polifenols previously divided into quintiles.

```{r}
names(main_db)[460:487]
PP_subclass_q5 <- names(main_db)[460:487]
#PP_subclass_q5_db <- main_db[,460:487]
length(PP_subclass_q5)

```
For some reason there are some differences between the subclasses of one set and the other:

```{r}
# gsub(".*n5_", "", PP_subclass_q5) #Removes n5_ from all occurences in string
setdiff(PP_subclass, gsub(".*n5_", "", PP_subclass_q5))

```
There are 13 differences between subclasses. Since there are only 9 more subclasses without quantile, this may be because a spelling error:

```{r}
setdiff(gsub(".*n5_", "", PP_subclass_q5), PP_subclass )
wrong_names <- setdiff(gsub(".*n5_", "", PP_subclass_q5), PP_subclass )

```
As previously suggested there are 4 quantized subgroups misspelled without capital letter.
Let's correct this mistake:

```{r}
capwords <- function(s, strict = FALSE) {
    cap <- function(s) paste(toupper(substring(s, 1, 1)),
                  {s <- substring(s, 2); if(strict) tolower(s) else s},
                             sep = "", collapse = " " )
    sapply(strsplit(s, split = " "), cap, USE.NAMES = !is.null(names(s)))
}


names_to_correct <-paste0("n5_",wrong_names) 
names_to_correct
which(names(main_db) %in% names_to_correct)


names_to_correct_ok <- paste0("n5_", capwords(wrong_names))
names_to_correct_ok

names(main_db)[which(names(main_db) %in% names_to_correct)] <- names_to_correct_ok

```
Now we can check again which subgroups are transformed into quintiles.


```{r}
names(main_db)[460:487]
PP_subclass_q5 <- names(main_db)[460:487]
#PP_subclass_q5_db <- main_db[,460:487]
length(PP_subclass_q5)

```

```{r}
# gsub(".*n5_", "", PP_subclass_q5) #Removes n5_ from all occurences in string
setdiff(PP_subclass, gsub(".*n5_", "", PP_subclass_q5))

```
Still there are 2 more than there should be.

```{r}
setdiff(gsub(".*n5_", "", PP_subclass_q5), PP_subclass )
wrong_names <- setdiff(gsub(".*n5_", "", PP_subclass_q5), PP_subclass )

```
Here the problem is pp instead of PP in subclass and in class PP is missing let's recode it: 

```{r}
names_to_correct <-paste0("n5_",wrong_names) 
names_to_correct
which(names(main_db) %in% names_to_correct)


names_to_correct_ok <- c("n5_Other_PP_subclass","n5_Other_PP_class")
names_to_correct_ok
```


```{r}
names(main_db)[which(names(main_db) %in% names_to_correct)] <- names_to_correct_ok

```

Now let's check again:

```{r}
names(main_db)[460:487]
PP_subclass_q5 <- names(main_db)[460:487]
length(PP_subclass_q5)
setdiff(PP_subclass, gsub(".*n5_", "", PP_subclass_q5))

```
```{r}
setdiff(gsub(".*n5_", "", PP_subclass_q5), PP_subclass )
```
Now they are correctly named.
The string `PP_subclass_notq5` will contain those variables which are not transformed into quantiles:

```{r}
PP_subclass_notq5 <- setdiff(gsub(".*n5_", "", PP_subclass_q5), PP_subclass )
```


It is convenient to maintain the same criteria for nomenclature when deciding subclasses.
My personal recomendation is to ignore this pre-build classes and create a new quantization in case it is required.

```{r}


quintile <- function(x) cut(x,
                            breaks = unique(quantile(x, c(0, 0.20, 0.4, 0.6, 0.8, 1))),
                            ordered_result = TRUE,
                            right=FALSE,
                            include.lowest = T,
                            labels =F,
                            type=3
)
PP_quintile <- data.frame(apply(main_db[,PP_subclass],2,quintile))
str(PP_quintile)


                 
#PP_subclass_q5_db <- gtools::quantcut(main_db$Anthocyanins,5)
```

```{r}
quantile(main_db$Theaflavins, c(0, 0.20, 0.4, 0.6, 0.8, 1))
```

There are some variables that have repeated quantile values. This results in repeted value for cuts, which makes fewer cuts. The solution is to increase the number of groups until 5 groups are formed. This is what the function cut2 makes. Allowing the same number of groups for all the variables.

```{r}
factor(Hmisc::cut2(main_db$Anthocyanins, g=5,m=5,minmax=T,include.lowest = TRUE))
```


```{r}
#factor(Hmisc::cut2(main_db$Theaflavins, g = 4), labels = c(1:4))

quintile <- function(x) factor(Hmisc::cut2(x,g=5,exclude=0,labels=c(1:4)))

PP_quintile <- data.table(apply(main_db[,PP_subclass],2,quintile))
#str(PP_quintile)
```
```{r}
library(data.table)
dtsetDT(PP_quintile)
PP_quintile[.SD , quartile := floor( 1 + 4 * (.I-1) / .N)]
```



6.Separate all polifenols:

```{r}
polifenoles <- datos[,4:422]
```

We are going to normalize the polifenols using different transformations

 1.- Logaritmic transformation:
Logaritms need a gamma coefficient in order to deal with 0. In order to choose this value we are going to check wich is the smallest value in all the dataset


```{r}
min(polifenoles[polifenoles != 0])
which.min(polifenoles[polifenoles != 0],polifenoles != 0)
```


## Data preparation:

